Identifying 15-minute neighbourhoods with complete caring potential

### 2024-10-08 paper/Manuscript.qmd
Draft for full team review

```{r library-setup}
library(biscale)
library(cowplot)
library(magick)
library(dplyr)
library(flextable)
library(ftExtra)
library(glue)
library(here)
library(mapview)
library(pander)
library(scales)
library(sf)
library(tmap)
library(tmaptools)
library(gt)
#install.packages("gtExtras")
library(gtExtras)
```

@@

@@

`r word_comment("I am only vaguely familar with this literature as of these months - how does this sound?", " ")`
In addition, classification algorithms have been useful tools in the identification of mobility and spatial typologies in relation to transportation planning. Machine learning approaches have often been deployed. For instance, @burkeGeospatialAnalysisFramework2022 uses spatially constrained multivariate clustering, a regionalisation technique, to develop urban form typologies related to the 15-Minute City. The k-means algorithm is another popular method; @ganUnderstandingUrbanMobility2020 applies it to analyse travel behaviour and classify metro stations based on mobility patterns. Similarly, [@montero2023role; @montero2023applying] cluster travel behavior profile typologies representing different population segments and place of residence in Barcelona Metropolitan Area. Combing classification techniques is also common; for instance, @delmelleDifferentiatingPathwaysNeighborhood2017 uses self-organizing maps (SOM) algorithm to group US neighbourhoods by minimizing dissimilarity in attributes before applying k-means for further classification. Similarly, @victorianoTimeSpaceMoney2020 uses SOM to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the decision tree algorithm to partition the data into interpretable classifications.

@@

```{r plot-figure2, out.width=600, out.height=600}
#| label: fig-Fig2
#| fig-cap: "Socio-economic and demographic variables that characterise accessibility to care destinations retrieved from the 2021 Canadian Census. All scales in quartiles. Basemap shapefiles are sourced from the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]."
knitr::include_graphics(glue::glue(here::here(), "/figures/pop_descriptive_plots_2.png"))
```

@@

@@

For the SOM step, the SOM implementation in the function 'trainSOM()' from {SOMbrero} R package is used [@VillaVialaneix2017]. The input variables include the 143,882 parcels, each as individual observations along with 15 variables: the 14 calculated accessibility scores $S_i^t$, normalized to the min-max range score within each $t$, and one diversity value $D_i$. Otherwise, defaults for all other parameters are assumed, relying on the data-driven heuristics embedded in the 'trainSOM()' function. Consequently, a 100 node (10 by 10 grid) SOM structure using euclidean distance and square topology is produced. Simply put, the SOM algorithm proceeds as follows: a 2D grid of nodes is created (in our case 100 nodes) as specified by the analyst, where each node will represent a point in the reduced-dimensional space. Upon initialization, each node is assigned a random weight vector of the same dimension as the input data (in our case, 15). From the input data, a random observation with its associated weight vector (i.e., one parcel point with 15 variables) is selected and the Euclidean distance between its weight vector and all nodes in the grid is calculated. The node with the smallest distance (i.e., the smallest dissimilarity) is labelled the 'best matching unit' as it is the node that best represents the input observation. After this best matching unit is identified, its own weight and its neighbouring nodes are updated to become more like the input observation. The process of finding best matching units and updating their weights is repeated for every observation, multiple times, until the results converge. As mentioned, this competitive learning process produces a 100 node SOM structure where each observation (parcel) is assigned to 1 node with an associated dissimilarity index. The SOM output is typically examined through a dissimilarity dendogram and an associated dissimilarity variance explained plot to select an appropriately representative 'superclusters' [@VillaVialaneix2017; @victorianoTimeSpaceMoney2020].

@@

## Overview of access to care destinations
How many care destinations can be reached within a 15-minute walk is summed for each of the four care categories defined in @bl-tbl1. The median parcel value for each DA is visualised in @fig-Fig4. Furthermore, the median parcel diversity of care destinations accessible is presented in @fig-Fig5. Together, @fig-Fig4 and @fig-Fig5 represent the 15 variables that served as inputs into the SOM algorithm.

@@

## Identification of (in)completely (un)caring neighbourhoods
<!--

A+ is now A+
A is now A
A- is now A-

B+ is now B+
B is now B
D is now B-

F is now D

-->

Notably, in @fig-Fig4, the concentration of opportunities for all category types is centered in the downtown of Hamilton, with the highest concentration being in the lower city (near the lake shore) and pockets of the high concentration further from the shoreline. Grocery-centric destinations appear to be the most concentrated, followed by health-centric and dependent-centric caring destinations. Errand-centric destinations are the most sprawled. In many ways, @fig-Fig4 mirrors the spatial distribution of care destinations (@fig-Fig3) as the range of 15-minutes of travel is small. In @fig-Fig5, areas that have high care accessibility tend to have high diversity as well, though there are exceptions in pockets of the city outside the downtown core that have lower but equally low levels of accessibility for all care categories.

Based on the SOM classification methodology discussed and the 15 variables represented in @fig-Fig4 and @fig-Fig5, 7 superclusters are identified from the 10 by 10 grid of SOM nodes. As diagnostics for the selection of the number of superclusters, the dissimilarity-index-based dendrogram with the proportion of parcels represented in each supercluster alongside the variance explained plot is visualised in @fig-Fig6.

```{r plot-figure6, out.width=800, out.height=800}
#| label: fig-Fig6
#| fig-cap: "The 7 resulting superclusters from the SOM the 10 by 10 grid output represented in a dendrogram (left) and the proportion of variance unexplained by cluster (right)."
#|
knitr::include_graphics(glue::glue(here::here(), "/figures/all_access_sc_plot_small.png"))
```
As seen in @fig-Fig6, the 7 superclusters are assigned labels from A to F, corresponding to their 'grade' of quantity of accessibility and diversity in accessibility per care category. What differentiates the dissimilarities in the clusters the most, i.e., nodes of grade A/A+ to everything else, is their high levels of care accessibility and often high diversity in care categories calculated for each parcel. A/A+ grades represent only 24% of the parcels, with the other grades being assigned lower accessibility and diversity of accessibility scores. To characterise the superclusters, a table providing summary statistics of each supercluster is detailed in @tbl-tbl2 and boxplots in @fig-Fig7.

```{r creating-table2-pt2}
#| tbl-cap: "Mean and (standard deviation) of each SOM classified cluster by input variable and additional summary variables (indicated by ALL CAPITAL LETERS)."
#| label: tbl-Tbl2
#|
summary_df %>%
  gt() %>%
  tab_header(
    title = " ",
  ) %>%
  cols_label(
    name = " "
  ) %>%
  gt_highlight_cols(
    2:3, fill = "green", alpha = 0.3
    ) %>%
  gt_highlight_cols(
    4, fill = "green", alpha = 0.2
    ) %>%
  gt_highlight_cols(
    5:6, fill = "orange", alpha = 0.3
    ) %>%
    gt_highlight_cols(
    7, fill = "orange", alpha = 0.2
    ) %>%
  gt_highlight_cols(
    8, fill = "red", alpha = 0.3
    ) %>%
  tab_style(
    style = cell_borders(
      sides = c("top"),
      color = "black",
      weight = px(1.5),
      style = "solid"
    ),
    locations = cells_body(rows = c(1,4,11,15,19))
  ) |>
  tab_options(
    table.font.size = 12,
    table.width = pct(80)
  )
```

@@

The superclusters are characterized by caring access and diversity of caring access (completeness). This technique provides a data-driven approach to supercluster formation which we then attribute meaning to through the derivation of the following labels. From the descriptive statistics of the superclusters, interesting trends emerge:

- A+, A, and A- all have exceptionally high or excellent caring accessibility for all destination types; all destination type and diversity scores are top quantile or above. Interestingly, A- is like A but with much higher dependent centric destinations, particularly parks, schools, and daycares, with high scores for everything else. This disbalance lowers its overall completeness score but is a supercluster that demonstrates potential in being retrofitted to provide A+ level completely caring access.  Together, these grades represent 28% of all parcels.
- B+, B and B- superclusters present about average completely caring access. These superclusters, particuarly B, can serve as the benchmark for what 'average' completely caring access in Hamilton currently looks like. B+ access and diversity scores are above B, while B- has low values for everything with exception to above average number of dependent-centric destinations, particularly parks, daycares and schools. These three clusters represent 34% of the parcels.
- D superclusters demonstrate the lows scores all around, presenting plenty of room for improvement in complete and caring 15-minute access. This cluster represents the 38% of all parcels.

@@

To meaningfully visualise where the superclusters are located, we visualise the representative supercluster label in each DA in @fig-Fig8. This is done by grouping parcels by their DA, calculating the median supercluster label for each label, and selecting the label which is most dominate within that DA. For reference, the median number of parcels in each DA is `r df_DA$parcel_count |> median() |> round()`. Further, supercluster label membership within a DA is typically pure, with the median membership being `r df_DA$max_label_val |> median() |> scales::percent()` of a single cluster.

`r word_comment("What do you think about these grades as labels? Maybe there's another way we can frame it. A+ and A are excellent, B+ and B are average, D are bad. However A- and B- show promise: they have some caring access (child-centric) and can be improved! A- can be more easily improved than B-.", " ")`
```{r plot-figure8, out.width=600, out.height=600}
#| label: fig-Fig8
#| fig-cap: "The maximum median parcel supercluster membership per DA. Escarpment is visualised as a grey dashed line."
knitr::include_graphics(glue::glue(here::here(), "/figures/SClabel_plot.png"))
```
The resulting superclusters visualised in @fig-Fig8 are a combination of the cumulative opportunity and diversity of access measures trasnlated into meaningful typologies through the SOM methodology. It can be observed that the excellent and exceptional neighbourhoods, which are very caring and very complete 15-minute neighbourhoods, are located within the center of the city and most closer to the shoreline in the 'lower' city (below the escarpment). The average and below average neighbourhoods are often proximate to the excellent and exceptional neighbourhoods and are more prevalent in the 'mountain' part of the center of Hamilton. The escarpment is a physical barrier, with few pedestrian-accessible access points to traverse; hence the typologies describing neighbouring DAs separated by the escarpment are often different.

As accessibility is the measure of _potential_ interaction, e.g., how many care destinations one could reach within a 15-minute walk, these typologies are useful in identifying which areas of the city are providing regionally high levels of complete caring access, which areas are regionally average (oranges), and which areas are regionally below average. Furthermore, neighbourhoods which provide disbalanced access can be further investigated. For instance, neighbourhoods that provide A- provide otherwise high levels of 15-minute walking access to all types of destinations, but particularly to Community and recreational centres, daycares, parks, and schools. These A- neighbourhoods are suitable for all, but may be particularly desirable for households that have children, needed proximity to care destinations. Furthermore, B- neighbourhoods are like A- neighbourhoods as they provide high access to children-centric destinations like schools, parks, and daycares, however are low in all other destination types. These neighbourhoods can be understood as currently being conducive to completing children-centric care trips, and having the potential to become complete if additional destination types were located within their neighbourhoods.

@@

To enhance the meaning of the superclusters beyond a descriptive and spatial conceptualisation of "caring" and "complete",  _who?_ resides in what neighbourhood currently can also be investigated. We run a decision tree and include the population-weighted socio-demographic characteristics of the parcels (retrieved from the 2021 Canadian Census) as input variables and the resulting output are 'socio-demographic superclusters' of the superclusters. The input variables included: median household income, % below the median household income, % LICO prevalence, average number of children per household, % population aged 0 to 14, % not in the labour force, % not employed, Gini index on adjusted household after-tax income, % visible minority, % single parent household, % who walk to work (relative to bike, care/truck/van, public transit and 'other'), % of owner household in core housing need (i.e., inadequate housing structure or paying higher than 30% income on housing), % of tenant households in subsidized housing, % of tenant households in core housing need, % no certification or with only a highschool diploma.

Of all the included input variables, the one that proves most meaningful in partitioning the superclusters data into subsets was only one variable: median household income of the DA. @fig-Fig9 provides the splits that were significant in partitioning the data in the decision tree algorithim. While the algorithim is unable to homogenously use each supercluster, we still use it to report a narrative of who may reside in what caring/complete supercluster. Particularly, the decision tree demonstrates a more pure supercluster membership for only two of the three terminal decision tree nodes: supercluster A+ (exceptional completely caring access) and D (not caring and complete access). For reference, the city-wide mean household income is `r df_DA_nogeom$Income |> mean(na.rm=T)|> round() |> prettyNum(big.mark = ",") ` (SD: `r df_DA_nogeom$Income |> sd(na.rm=T)|> round() |> prettyNum(big.mark = ",")` and median: `r quantile(df_DA_nogeom$Income,0.5, na.rm=T) |> unname()  |> prettyNum(big.mark = ",")`).

@@

```{r plot-figure9, out.width=600, out.height=300}
#| label: fig-Fig9
#| fig-cap: "The decision tree demonstrating socio-ecnomic profiles of superclusters."
knitr::include_graphics(glue::glue(here::here(), "/figures/access_sc_profiles_justincome.png"))
```
Describing @fig-Fig9, the most valuable variable for splitting the data was the median household income in each DA. The terminal nodes represented as bar columns depict the percentage composition by supercluster. For the column on the right, the majority of the parcels are D superclusters with the second most supercluster membership being B- parcels, described by having greater than 91,500 CAD household income, above the city-wide mean.  The left column is dominated by A+ along with A and A- (in the largest quantities relative to other bars), though proportions of all other superclusters are also present. This column is defined by parcels present in DAs with lower incomes, e.g., 68,750 CAD. The middle columns's classification is mixed: with all types of superclusters present in DAs with about average households incomes (between 68,750 CAD and 91,500 CAD).

Though median income was the most useful in partitioning the parcels by their supercluster labels, other variables are still important but were not as helpful in partitioning the data. Namely, the following variables were also important, in order of importance and correlation coefficient with the median household income variable in brackets: the % below the median household income (`r Income_inspect[c("Income_bottom"),] |> round(digits=2)`), % single parent household (`r Income_inspect[c("One_parent_families"),] |> round(digits=2)`), % no certification or with only a highschool diploma (`r Income_inspect[c("Low_or_no_deg"),] |> round(digits=2)`), the average number of children per household (`r Income_inspect[c("Avg_numb_of_kids"),] |> round(digits=2)`), % not in the labour force (`r Income_inspect[c("Not_in_labforce"),] |> round(digits=2)`), % LICO prevalence (`r Income_inspect[c("LICO"),] |> round(digits=2)`), and  % who walk to work (`r Income_inspect[c("walk_towork"),] |> round(digits=2)`). Of note, when median household income is removed from the decision tree, a similar structure is created, except with above average Single-Parent family and LICO prevalence splitting the data. It is notable that median household income is highly or moderately correlated with many of these important variables, and discussing these correlation is useful for interpretation alongside the tree diagram.

For instance, single-parent families are most positively correlated with the proportion of households being in the bottom income distribution (`r Income_inspect[c("Income_bottom"),] |> round(digits=2)`) and low or new degree (`r One_parent_family_inspect[c("Low_or_no"),] |> round(digits=2)`), and most negatively with the median after-tax household income (`r One_parent_family_inspect[c("v_CA21_907: Median after-tax income of household in 2020 ($)"),] |> round(digits=2)`), reflecting nationwide trends [@governmentofcanadaDailyMainHighlights2024; @governmentofcanadaPrevalenceLowIncome2024]. LICO is also an indicator of low-income, and though not all single-family households are below the LICO (in fact, single-family households are not highly correlated with LICO at the DA level in Hamilton), DAs where higher concentrations of single-family and where high LICO prevalence persist tend to have A+ access. These findings are notable as currently, the highest grade (A+) serves a disadvantaged group that is also often disadvantaged along different intersecting socio-economic characteristics. Also interestingly, DAs with lower single-parent households (who also tend to have higher incomes) are in DAs with a higher concentration of F parcels. These considerations are important when considering the equity of caring 15-minute policy interventions.

<!-- Describing @fig-Fig9, the most valuable variable for splitting the data was the proportion of single family households and % of LICO prevalence in the DA. The the terminal nodes represented as bar columns depict the percentage composition by supercluster. For the bar on the right, the majority of the parcels are F superclusters. Following the tree diagram, these F superclusters are described by having less than 15.8% single parent families (this is below the median of `r df_DA$One_parent_families |> median(na.rm=T) |> scales::percent()`). The bar on the left is mostly composed of parcels that are labeled A+. These A+ parcels are located in DAs greater than 15.8% single parent families and % LICO prevalence of 6.5% and greater (this is above the median of `r df_DA$LICO |> median(na.rm=T) |> round() |> paste0("%")`). In sum: residences in parcels with exceptionally complete and caring access tend to be lower-income and single-parent families, while residences in parcels with not caring are not complete access tend to have a lower presence of single parent households. Households with a higher proportion of single-family households (15.8% and greater) where LICO is lower have a mixed label membership (bar column in @fig-Fig9).  -->

@@

The descriptive statistics associated with each supercluster, their spatial distribution, and the socio-economic profiles that characterise them, are revealing points of discussions and further examined.

>>>>

```{r library-setup}
library(biscale)
library(cowplot)
library(magick)
library(dplyr)
library(flextable)
library(ftExtra)
library(glue)
library(here)
library(mapview)
library(pander)
library(scales)
library(sf)
library(tidyr)
library(tmap)
library(tmaptools)
library(gt)
#install.packages("gtExtras")
library(gtExtras)
library(tidyr)
```

@@

```{r population in each bar column for the Decision Tree, eval=FALSE}
df_DA |> filter(`v_CA21_907: Median after-tax income of household in 2020 ($)` <= 68750) |>  select("Population") |> st_drop_geometry()  |> sum(na.rm=T) / df_DA |>  select("Population") |> st_drop_geometry()  |> sum(na.rm=T)

df_DA |> filter(`v_CA21_907: Median after-tax income of household in 2020 ($)` > 68750 & `v_CA21_907: Median after-tax income of household in 2020 ($)` <  91500) |>  select("Population") |> st_drop_geometry()  |> sum(na.rm=T) / df_DA |>  select("Population") |> st_drop_geometry()  |> sum(na.rm=T)

df_DA |> filter(`v_CA21_907: Median after-tax income of household in 2020 ($)` >=  91500) |>  select("Population") |> st_drop_geometry()  |> sum(na.rm=T) / df_DA |>  select("Population") |> st_drop_geometry()  |> sum(na.rm=T)
```

@@

In addition, classification algorithms have been useful tools in the identification of mobility and spatial typologies in relation to transportation planning. Machine learning approaches have often been deployed. For instance, @burkeGeospatialAnalysisFramework2022 uses spatially constrained multivariate clustering, a regionalisation technique, to develop urban form typologies related to the 15-Minute City. The k-means algorithm is another popular method; @ganUnderstandingUrbanMobility2020 applies it to analyse travel behaviour and classify metro stations based on mobility patterns. Similarly, [@montero2023role; @montero2023applying] cluster travel behavior profile typologies representing different population segments and place of residence in Barcelona Metropolitan Area. Combing classification techniques is also common; for instance, @delmelleDifferentiatingPathwaysNeighborhood2017 uses self-organizing maps (SOM) algorithm to group US neighbourhoods by minimizing dissimilarity in attributes before applying k-means for further classification. Similarly, @victorianoTimeSpaceMoney2020 uses SOM to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the decision tree algorithm to partition the data into interpretable classifications.

@@

`r word_comment("After reading the results and discussion, do you think fig 2 plots are helpful? Maybe they should be removed?", " ")`
```{r plot-figure2, out.width=600, out.height=600}
#| label: fig-Fig2
#| fig-cap: "Socio-economic and demographic variables that characterise accessibility to care destinations retrieved from the 2021 Canadian Census. All scales in quartiles. Basemap shapefiles are sourced from the Open Data Hamilton Portal [@opendatahamiltonCityBoundary2023] and the USGS [@greatlakesUSGS2010]."
knitr::include_graphics(glue::glue(here::here(), "/figures/pop_descriptive_plots_2.png"))
```

@@

`r word_comment("Any ideas on how to condense this table...? Maybe we can cite the forthcoming JTG article, what do you think?", " ")`

@@

For the SOM step, the SOM implementation in the function 'trainSOM()' from {SOMbrero} R package is used [@VillaVialaneix2017]. The input variables include the 143,882 parcels, each as individual observations along with 15 variables: the 14 calculated accessibility scores $S_i^t$, normalized to the min-max range score within each $t$, and one diversity value $D_i$. Otherwise, defaults for all other parameters are assumed, relying on the data-driven heuristics embedded in the 'trainSOM()' function. Consequently, a 100 node (10 by 10 grid) SOM structure using euclidean distance and square topology is produced. Simply put, the SOM algorithm proceeds as follows: a 2D grid of nodes is created as specified by the analyst, where each node will represent a point in the reduced-dimensional space. Upon initialization, each node is assigned a random weight vector of the same dimension as the input data (in our case, 15). From the input data, a random observation with its associated weight vector (i.e., one parcel point with 15 variables) is selected and the Euclidean distance between its weight vector and all nodes in the grid is calculated. The node with the smallest distance (i.e., the smallest dissimilarity) is labelled the 'best matching unit' as it is the node that best represents the input observation. After this best matching unit is identified, its own weight and its neighbouring nodes are updated to become more like the input observation. The process of finding best matching units and updating their weights is repeated for every observation, multiple times, until the results converge. As mentioned, this competitive learning process produces a 100 node SOM structure where each observation (parcel) is assigned to 1 node with an associated dissimilarity index. The SOM output is typically examined through a dissimilarity dendogram and an associated dissimilarity variance explained plot to select an appropriately representative 'superclusters' [@VillaVialaneix2017; @victorianoTimeSpaceMoney2020].

@@

## Overview of access to care destinations
How many care destinations can be reached within a 15-minute walk is summed for each of the four care categories defined in @tbl-Tbl1. The median parcel value for each DA is visualised in @fig-Fig4 and the median parcel diversity of care destinations accessible is presented in @fig-Fig5. Together, @fig-Fig4 and @fig-Fig5 represent the 15 variables that served as inputs into the SOM algorithm.

@@

## Identification of (in)completely (un)caring neighbourhoods

In @fig-Fig4, the concentration of opportunities for all category types is centered in the downtown of Hamilton, with the highest concentration being in the lower city (near the lake shore) and pockets of the high concentration further from the shoreline. Grocery-centric destinations appear to be the most concentrated, followed by health-centric and dependent-centric caring destinations. Errand-centric destinations are the most sprawled. In many ways, @fig-Fig4 mirrors the spatial distribution of care destinations (@fig-Fig3) as the range of 15-minutes of travel is small. In @fig-Fig5, areas that have high care accessibility tend to have high diversity as well, though there are exceptions in pockets of the city outside the downtown core that have lower but equally low levels of accessibility for all care categories.

Based on the SOM methodology discussed, 7 superclusters are identified from the 10 by 10 grid of SOM nodes. As diagnostics for the selection of the number of superclusters, the dissimilarity-index-based dendrogram with the proportion of parcels represented in each supercluster alongside the variance explained plot is visualised in @fig-Fig6. Labels representing grades A to D qualifying the 7 superclusters are then assigned. These grades as also reflected in @fig-Fig6 and represent the quantity of accessibility and diversity in accessibility per care category. Higher grades (A+ and A) corresponding to the highest accessibility and diversity, while lowest grades (D) representing lowest accessibility and diversity scores.

```{r plot-figure6, out.width=800, out.height=800}
#| label: fig-Fig6
#| fig-cap: "The 7 resulting superclusters from the SOM the 10 by 10 grid output represented in a dendrogram (left) and the proportion of variance unexplained by cluster (right)."
#|
knitr::include_graphics(glue::glue(here::here(), "/figures/all_access_sc_plot_small.png"))
```

The grade for each supercluster was carefully assigned by the authors by referring to descriptive statistics and their interesting trends. These statistics are summarised in @tbl-Tbl2 and in the boxplots in @fig-Fig7.

- A+ and A have exceptionally high caring accessibility for all destination types and high diversity scores in the top quantile or above. Together, these grades represent 24% of all parcels.
- A- has high caring accessibility, but low diversity. Interestingly, A- is like A but with much higher dependent-centric destinations, particularly parks, schools, and daycares, with only moderately high scores for all  other destination types. This disbalance lowers its overall diversity score. However, it can be characterized as a supercluster that demonstrates potential in being retrofitted to provide A+ level completely caring access. This grade  represent 4% of parcels. 
- B+ and B superclusters present about average completely caring access. These superclusters can serve as the benchmark for what 'average' completely caring access in Hamilton currently looks like and they represent 15% of parcels.
- B- provides above average dependent-centric destination access, particularly parks, daycares and schools, but below average access to other destination types and hence has low diversity scores. B- is like A- as it demonstrates complete caring 15-minute potential if retrofitted. As these parcels demonstrate caring access to some destinations, they may have the potential to be retrofitted to support complete access to all caring destination types. This grade represents 19% of parcels.
- D superclusters demonstrate the lowest scores all-around, representing room for land-use improvement that addressed complete and caring 15-minute access. This supercluster characterizes the most number of parcels, representing 38% of parcels in the city.

```{r creating-table2-pt2}
#| tbl-cap: "Mean and (standard deviation) of each SOM classified cluster by input variable and additional summary variables. Variables included in the SOM algorthim are in regular case, while additional summary variables are indicated by ALL CAPITAL LETERS."
#| label: tbl-Tbl2
#|
# summary_df %>%
#   gt() %>%
#   tab_header(
#     title = " ",
#   ) %>%
#   cols_label(
#     name = " "
#   ) %>%
#   gt_highlight_cols(
#     2:3, fill = "green", alpha = 0.3
#     ) %>%
#   gt_highlight_cols(
#     4, fill = "green", alpha = 0.2
#     ) %>%
#   gt_highlight_cols(
#     5:6, fill = "orange", alpha = 0.3
#     ) %>%
#     gt_highlight_cols(
#     7, fill = "orange", alpha = 0.2
#     ) %>%
#   gt_highlight_cols(
#     8, fill = "red", alpha = 0.3
#     ) %>%
#   tab_style(
#     style = cell_borders(
#       sides = c("top"),
#       color = "black",
#       weight = px(1.5),
#       style = "solid"
#     ),
#     locations = cells_body(rows = c(1,4,11,15,19))
#   ) |>
#   tab_options(
#     table.font.size = 10,
#     table.layout = 'fixed',
#     table.align = 'center',
#     table.width = pct(100)
#  )

set_flextable_defaults(font.size= 8,
                       font.family = NULL,
                       text.align = 'justify',
                       table.layout = 'fixed')

summary_df <- summary_df |> rename(" " = "name")

flextable(summary_df) %>%
 ftExtra::colformat_md() %>%
  flextable::width(j = 1, width = 3.06, unit = "cm") %>%
  flextable::width(j = 2:9, width = 1.53, unit = "cm") %>%
  bg(j= 2, bg = "#48B376") |>
  bg(j= 3, bg = "#91CF60") |>
  bg(j= 4, bg = "#D9EF8B") |>
  bg(j= 5, bg = "#FFFFBF") |>
  bg(j= 6, bg = "#FEE08B") |>
  bg(j= 7, bg = "#FDAE61") |> 
  bg(j= 8, bg = "#D75C56") |>
  hline(i = c(3,10,14,18))

# RColorBrewer::brewer.pal(7, "RdYlGn")
# "#D73027" "#FC8D59" "#FEE08B" "#FFFFBF" "#D9EF8B" "#91CF60" "#1A9850"
```

@@

@@

To spatially demonstrate where the superclusters are located, representative supercluster labels for each DA in visualised in @fig-Fig8. This visualisation is created by grouping parcels by their DA, calculating the median supercluster label for each label, and selecting the label that is most dominant within that DA. For reference, the median number of parcels in each DA is `r df_DA$parcel_count |> median() |> round()` and supercluster label membership within a DA is typically pure, with the median membership being `r df_DA$max_label_val |> median() |> scales::percent()` of a single cluster.

`r word_comment("What do you think about these grades as labels? Maybe there's another way we can frame it. A+ and A are excellent, B+ and B are average, D are bad. However A- and B- show promise: they have some caring access (child-centric) and can be improved! A- can be more easily improved than B-.", " ")`
```{r plot-figure8, out.width=600, out.height=600}
#| label: fig-Fig8
#| fig-cap: "The maximum median parcel supercluster membership per DA. Escarpment is visualised as a grey dashed line."
knitr::include_graphics(glue::glue(here::here(), "/figures/SClabel_plot.png"))
```
The resulting superclusters visualised in @fig-Fig8 are a combination of the cumulative opportunity and diversity of access measures translated into meaningful typologies through the SOM methodology. Accessibility is the measure of _potential_ interaction, e.g., how many care destinations one could reach within a 15-minute walk, these typologies are useful in identifying which areas of the city are providing regionally high levels of complete caring access, which areas are regionally average, and which areas are regionally below average. Furthermore, neighbourhoods which provide disbalanced access can be further investigated.

In @fig-Fig8, it can be observed that the excellent and exceptional neighbourhoods (A, A+), which are very caring and very complete 15-minute neighbourhoods, are located within the center of the city and closest to the shoreline in the Lower City. The above and below average neighbourhoods (B+ B) are often proximate to the excellent and exceptional neighbourhoods and are more prevalent in the Mountain part of the center of Hamilton. The escarpment is a physical barrier, with few pedestrian-accessible access points to traverse; hence the typologies describing neighbouring DAs separated by the escarpment are often different. Below average (D) neighbourhoods are located in peripheral areas outside the center of Hamilton, in areas where urban form is characterised by lower density residential housing and auto-dependent mobility.

Furthermore in @fig-Fig8, A- and B- grades stand out as offering high access for children-centric destinations for their grade-group but below group-average access in other destination types. These neighbourhoods may be more suitable for populations who prioritize walkable access to destinations like schools, parks and daycares, and find access to other types of caring destinations less important. These neighbourhoods also stand out as demonstrating potential to be retrofitted to provide more complete access if additional destination types were located within their neighbourhoods.

@@

To enhance the meaning of the superclusters beyond a descriptive and spatial conceptualisation of "caring" and "complete",  _who?_ resides in what neighbourhood is investigated through the Decision Tree results. The input features of the Decision Tree are the supercluster labels and the feature variables are various population-weighted socio-demographic characteristics of the 2021 Canadian Census, namely: median household income, % below the median household income, % LICO prevalence, average number of children per household, % population aged 0 to 14, % not in the labour force, % not employed, Gini index on adjusted household after-tax income, % visible minority, % single parent household, % who walk to work (relative to bike, care/truck/van, public transit and 'other'), % of owner household in core housing need (i.e., inadequate housing structure or paying higher than 30% income on housing), % of tenant households in subsidized housing, % of tenant households in core housing need, % no certification or with only a highschool diploma.

Of all the included input variables, _median household income_ proved to be the most meaningful in partitioning the superclusters data. @fig-Fig9 provides the Decision Tree with the significant splits in median household income and the proportion composition of supercluster along each branch for three terminal Decision Tree nodes. While the algorithm is unable to homogenously use each supercluster, @fig-Fig9 is helpful to report a narrative of who may reside in what caring/complete supercluster. Particularly, the Decision Tree demonstrates a more pure supercluster membership for only two of the three terminal decision tree nodes: supercluster A+ (exceptional completely caring access) and D (not caring and complete access). For reference, the city-wide mean household income is `r df_DA_nogeom$Income |> mean(na.rm=T)|> round() |> prettyNum(big.mark = ",") ` (SD: `r df_DA_nogeom$Income |> sd(na.rm=T)|> round() |> prettyNum(big.mark = ",")` and median: `r quantile(df_DA_nogeom$Income,0.5, na.rm=T) |> unname()  |> prettyNum(big.mark = ",")`). The notable splits for income are $\gt$ \$91,500 $\lt$ \$68,750, or between \$68,750 and \$91,500, roughly split by lower income, middle income, and higher income median household brackets.

@@

```{r plot-figure9, out.width=600, out.height=300}
#| label: fig-Fig9
#| fig-cap: "The decision tree demonstrating median household income splits and superclusters composition within each branch."
knitr::include_graphics(glue::glue(here::here(), "/figures/access_sc_profiles_justincome.png"))
```
Describing @fig-Fig9, the bar column on the right represents DAs with higher household income. In these DAs, the majority of parcels are labeled as D superclusters, i.e not caring or complete 15-minute neighbourhoods. Within this higher-income-representing bar column, the second highest proportion of supercluster membership are B- parcels. B- parcels are low in most amenity types except a few child-centric destinations and show promise in being more easily improved than the D parcels. The higher income column represents 39% of the population, the largest proportion of any of the three columns. The left column in @fig-Fig9 correpnds to the lowest income, i.e., below 68,750 CAD median household income, and accounts for 32% the population. It is dominated by A+ parcels along with A and A- parcels, in the largest quantities relative to other bar columns, however, proportions of all other superclusters are also present. The middle column is defined by parcels with a mix of supercluster classifications and represents DAs with middle households incomes, between 68,750 CAD and 91,500 CAD representing 28% of the population.

Though median income was the most useful in partitioning the parcels by their supercluster labels in @fig-Fig9, other variables are still important. Namely, the following variables were also important (in order of importance and correlation coefficient with the median household income variable in brackets): the % below the median household income (`r Income_inspect[c("Income_bottom"),] |> round(digits=2)`), % single parent household (`r Income_inspect[c("One_parent_families"),] |> round(digits=2)`), % no certification or with only a highschool diploma (`r Income_inspect[c("Low_or_no_deg"),] |> round(digits=2)`), the average number of children per household (`r Income_inspect[c("Avg_numb_of_kids"),] |> round(digits=2)`), % not in the labour force (`r Income_inspect[c("Not_in_labforce"),] |> round(digits=2)`), % LICO prevalence (`r Income_inspect[c("LICO"),] |> round(digits=2)`), and % who walk to work (`r Income_inspect[c("walk_towork"),] |> round(digits=2)`). <!-- Of note, when median household income is removed from the decision tree, a similar structure is created, except with above average Single-Parent family and LICO prevalence splitting the data.-->

It is notable that median household income is highly or moderately correlated with many of these important variables, and discussing these correlation is useful for interpretation alongside the Decision Tree diagram. For instance, single-parent families are most positively correlated with the proportion of households being in the bottom income distribution (`r Income_inspect[c("Income_bottom"),] |> round(digits=2)`) and low or new degree (`r One_parent_family_inspect[c("Low_or_no"),] |> round(digits=2)`), and most negatively with the median after-tax household income (`r One_parent_family_inspect[c("v_CA21_907: Median after-tax income of household in 2020 ($)"),] |> round(digits=2)`), reflecting nationwide trends [@governmentofcanadaDailyMainHighlights2024; @governmentofcanadaPrevalenceLowIncome2024]. Though not all single-family households are below the LICO (single-family households are not highly correlated with LICO at the DA-level in Hamilton), DAs with higher concentrations of single-family households and LICO prevalence tend to be A+ completely caring 15-minute neighbourhood. Conversely, DAs with lower single-parent households are in DAs with a higher concentration of F parcels. These findings are notable from an equity perspective and in considering caring 15-minute policy interventions, especially since economic disadvantaged often intersects with other socio-demographic characteristics.

@@

`r word_comment("Ideas on this section (idea for an extra paragraph or additional points to each paragraph to make them stronger, with citations) would greatly be appreciated!", " ")`

Geographically, Hamilton is city that offers very completely caring 15-Minute neighbourhoods for some, but not everyone. There are evident spatial inequalities in areas that are, are not, and are somewhere between a caring and complete 15-minute neighbourhood (from labels A+ and A being excellent, B being about average, D being well below average, and A- and B- demonstrating potential in their grade group). The downtown core of the city has the most _caring 15-minute neighbourhoods_, but many pockets of the suburbs do as well. Though this contrasts what may be seen when considering employment opportunities [as in the work of @elgeneidyNonstopEquityAssessing2016], the paper's findings align with the neighbourhood unit planning literature reviewed. As many North American Suburbs, Hamilton followed certain characteristic design patterns that were initially designed to be proximate to amenities, coinciding with the conceptualisation of _caring 15-minute cities_ [@aliNeighbourhoodPlanningHamilton1991].
